---
title: "Olympics EDA"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
  word_document: default
editor_options:
  chunk_output_type: console
---

In this file we are going to explore the data and find out the insights in our 
data.

```{r warning=FALSE, include=FALSE}
#install.packages("ggplot2")
#install.packages("dplyr")
#install.packages("gridExtra")
#install.packages("reshape2")
library(ggplot2)
library(dplyr)
library(gridExtra)
library(reshape2)
```

First of all we are going to load the data and look at the summary

```{r echo=FALSE, warning=FALSE}
data <- read.csv("data/data.csv")
summary(data)
```

Now that we have data. We are going to take a look at each column 
independently

## Univariant Analysis

### Sex

```{r echo=FALSE, warning=FALSE}
summary(data$Sex)
ggplot(aes(x=Sex), data=data)+
  geom_bar(fill='#555555')
```

Here we see that number of female atheletes are less than number of male 
athletes in olympic.

### Age

We know that there are missing values in our Age, Height and Weight Category
I am going to leave those values and plot the data that we have.

```{r echo=FALSE, warning=FALSE}
summary(data$Age)
ggplot(aes(x=Age), data=data) +
  geom_histogram(binwidth = 1, color='#ffffff') +
  scale_x_continuous(breaks = seq(0, 100, 10))
```

Based on the data available we can say that most of the athletes we have are 
young, with max altheletes of age 24

### Height

```{r echo=FALSE, warning=FALSE}
summary(data$Height)
ggplot(aes(x=Height), data=data) +
  geom_histogram(binwidth = 1, color='#ffffff') +
  scale_x_continuous(breaks = seq(0, 250, 10)) +
  xlab("Heights (cms)")
```

The distribution of Heights of atheletes is almost normal with some spikes.
As we can see these spiked values are either at 10s or 5s values These might 
be because of rounding off the values to nearest multiple of 5.

### Weight

```{r echo=FALSE, warning=FALSE}
summary(data$Weight)
ggplot(aes(x=Weight), data=data) +
  geom_histogram(binwidth = 1, color='#ffffff') +
  scale_x_continuous(breaks = seq(0, 220, 10))+
  xlab("Weights (Kgs)")
```
The max value for the weiht is 214, but most of the values lie between 
30 and 130. Again we see spikes at multiples of 10 and small spikes at multiples
of 5.

### Team

Team is the name of the team athelete belongs to.

```{r echo=FALSE, warning=FALSE}
length(unique(data$Team))
```

So we have 1184 total teams in olympics.

```{r echo=FALSE, warning=FALSE}
x<-sort(table(data$Team),decreasing=T)
print("Top 10 Team in terms of participation")
x[1:10]

y<- table(x)
y[1:10]

```

United states has max number of teams paticiaption followed by Frace and Great
Britain. But most of the teams has only single participation. We have 103 Teams 
with single participation, 265 teams with double paricipation and 168 with
triple.

### Sport

```{r echo=FALSE, warning=FALSE}
print("Number of sports")
length(unique(data$Sport))
x<-sort(table(data$Sport), decreasing = T)
x <- data %>% group_by(Sport) %>% summarise(count=n())
x$Sport <- factor(x$Sport, levels = x$Sport[order(-x$count)])
ggplot(aes(x=Sport, y=count), data=x)+
  geom_bar(width = 0.2, color='#005050', stat="identity")+
  scale_y_log10()+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))+
  ylab("count (log10 scaled)")

```

In this graph we can see the paticipation of athletes in Sports. Aeronautics has
the least number of participants ,only 1 and Athletics has maximum number of 
participants.


### Medals

We have three type of medals Gold, Silver and Bronze, If an athlete did not win
medal then DNW

I am going to plot only the Medal data when the athlete won

```{r echo=FALSE, warning=FALSE}
data$Won <- ifelse(data$Medal!='DNW', 1, 0)
data$Medal <- factor(data$Medal, levels = c('Gold', 'Silver', 'Bronze', 'DNW'))
sum(data$Won)
ggplot(aes(x=Won), data=data)+
  geom_bar(fill='#205050') 
summary(data$Medal)
```

The number of Gold, Silver and Bronze medals are not same, there is a difference 
of a few 100s.

## Bivariant Analysis

In this session we are going to take a look at relation between two variables

### Number of Male and Female athlete over the years

We are going to plot how the number of male and female athlete participation 
changed over the year

```{r echo=FALSE, warning=FALSE}
number_of_athletes<-data %>% 
  group_by(Year, Sex, Season) %>%
  summarise(number_of_athletes = length(unique(ID)))
p1<-ggplot(aes(x=Year, y=number_of_athletes), 
       data=subset(number_of_athletes, Season=='Summer'))+
  geom_line(aes(color=Sex))+
  geom_point(aes(color=Sex))+
  scale_x_continuous(breaks = seq(1896, 2016, 4))+
  ylab("Number of athletes")+
  ggtitle("Number of athletes in Summer Games")+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))
p2<- ggplot(aes(x=Year, y=number_of_athletes), 
       data=subset(number_of_athletes, Season=='Winter'))+
  geom_line(aes(color=Sex))+
  geom_point(aes(color=Sex))+
  ylab("Number of athletes")+
  scale_x_continuous(breaks = c(seq(1896, 1992, 4), seq(1994, 2016, 4)))+
  ggtitle("Number of athletes in Winter Games")+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))
grid.arrange(p1, p2, ncol=1)
```

In the first Olympics games we have no women athlete. But after that we see a 
grdual increae in women athlete It is very low in the begining and 
then we see a big rise in 80s. 

We also have some missin games 1916 summer, 1940-1944 summer and winter.
These are the times of World wars and the games were cancelled. 

Also there is a shoft in years of winter olympics. In 1994 the winter olympic 
year moved to 2 years after summer oympics. And before 1924 we didn't have 
winter olympics.

We see that the gap in number of male and female athletes in descresing over the
years. 

### Number of countries particiapted

```{r echo=FALSE, warning=FALSE}
participating_countries <- data %>%
                        group_by(Year, Season) %>%
                        summarise(Number_of_countries = length(unique(region)))
ggplot(aes(x=Year, y=Number_of_countries), data=participating_countries)+
  geom_line(aes(color=Season))+
  geom_point(aes(color=Season))+
  scale_x_continuous(breaks = seq(1896, 2016, 4))+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))+
  ylab("Number of Countries")

```

Over the years games saw an increse in number of countries. But in year 1980 we
see a shape decline in summer games. this is the same decline we saw in number 
of athletes too.

This is because the Soviet Union invaded Afghanistan and 1980 games were held in 
Soviet Union. United States initiated the boycot of games as a protest against 
invasion.

### countries and Medals

Now we are going to check if there are any countries that never won any medal 
at all, won other but never won gold medal, where either men or women never won gold medal.

```{r echo=FALSE, warning=FALSE}
medal_won <- data %>% group_by(region) %>%
                    summarise(number_of_medals = sum(Won))
no_medal <- subset(medal_won, number_of_medals==0)
print("Countries with no medal")
no_medal$region

all_medal_data <- data %>% group_by(region, Medal) %>%
                      summarise(total = n())
medal_data.wide <- dcast(all_medal_data, region ~ Medal)
medal_data.wide[is.na(medal_data.wide)] <- 0
no_gold_data <- subset(medal_data.wide, Gold==0 & Silver>0 & Bronze>0)
print("Countries with no Gold medal but others")
no_gold_data$region

all_medal_sex <- data%>% group_by(region, Medal, Sex) %>%
                         summarise(total = n())
all_medal_sex.wide <- dcast(all_medal_sex, region ~ Medal+Sex)
all_medal_sex.wide[is.na(all_medal_sex.wide)] <- 0
no_women_gold <- subset(all_medal_sex.wide, Gold_F ==0 & Gold_M>0 )
no_men_gold <- subset(all_medal_sex.wide, Gold_F>0 & Gold_M==0 )
print("countries where women never won gold medal but men has")
no_women_gold$region
print("countries where men never won gold medal but women has")
no_men_gold$region

```


So we have 72 nations that never won any medal. 11 countries never won a 
gold medal but they have won others. In 33 countries women never won Gold medals
but men have. In 8 countries men never won Gold medals but women have.


## Conclusion

1. Number of athletes incresed over the years. During the first year we did not 
had women athletes but then the numbers increased over the years

2. The number of countries participatin in Olympics also increased, but in year
1980 because of the boycott.

3. 72 nations never won any medal in Olympics

4. 11 countries never won gold medals but have won others

5. In 33 countries women have never won gold medals but men have 

6. In 8 countries men have never won gold medals but women have 


Now that we have these results from our Data Explorarion, I am going to save
medal data to seaprate files for the visualizations

```{r echo=FALSE, warning=FALSE}
write.csv(no_men_gold, "data/no_men_gold.csv", row.names = FALSE)
write.csv(no_women_gold, "data/no_women_gold.csv", row.names = FALSE)
write.csv(no_gold_data, "data/no_gold.csv", row.names = FALSE)
write.csv(no_medal, "data/no_medal.csv", row.names = FALSE)
```

