---
title: "Olympics EDA"
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

In this file we are going to explore the data and find out the insights in our 
data.

```{r echo=FALSE, warning=FALSE}

library(ggplot2)
#library(gridExtra)
library(dplyr)
#library(alr3)
#library(reshape2)
#library(GGally)
#library(scales)
#library(memisc)
#library(lattice)
#library(MASS)
#library(car)
#library(RColorBrewer)

```

First of all we are going to load the data

```{r echo=FALSE, warning=FALSE}
data <- read.csv("data/data.csv")
summary(data)
```

Now that we have data. First of all we going to take a look at each column 
independently

## Univariant Analysis

### Sex

```{r}
summary(data$Sex)
ggplot(aes(x=Sex), data=data)+
  geom_bar(fill='#555555')
```
Here we see that number of female atheletes are less than number of male 
athletes in olympic.

### Age

We know that there are missing values in our Age, Height and Weight Category
I am going to leave those values and plot the data that we have.

```{r}
summary(data$Age)
ggplot(aes(x=Age), data=data) +
  geom_histogram(binwidth = 1, color='#ffffff') +
  scale_x_continuous(breaks = seq(0, 100, 10))
```

Based on the data available we can say that most of the athletes we have are 
young, with max altheletes of age 24

### Height

```{r}
summary(data$Height)
ggplot(aes(x=Height), data=data) +
  geom_histogram(binwidth = 1, color='#ffffff') +
  scale_x_continuous(breaks = seq(0, 250, 10)) +
  xlab("Heights (cms)")
```

The distribution of Heights of atheletes is almost normal with some spikes.
As we can see these spiked values are either at 10s or 5s values These might 
be because of rounding off the values to nearest multiple of 5.

### Weight

```{r}
summary(data$Weight)
ggplot(aes(x=Weight), data=data) +
  geom_histogram(binwidth = 1, color='#ffffff') +
  scale_x_continuous(breaks = seq(0, 220, 10))+
  xlab("Weights (Kgs)")
```
The max value for the weiht is 214, but most of the values lie between 
30 and 130. Again we see spikes at multiples of 10 and small spikes at multiples
of 5.

### Team

Team is the name of the team athelete belongs to.

```{r}
length(unique(data$Team))
```

So we have 1184 total teams in olympics.

```{r}
x<-sort(table(data$Team),decreasing=T)
print("Top 10 Team in terms of participation")
x[1:10]

print("Bottom 50 Team")
tail(x, n=50)

y<- table(x)
y
```
United states has max number of teams paticiaption followed by Frace and Great
Britain. But most of the teams has only single participation. We have 103 Teams 
with single participation, 265 teams with double paricipation and 168 with
triple.

### Sport

```{r}
length(unique(data$Sport))
x<-sort(table(data$Sport), decreasing = T)
min(x)
max(x)
x <- data %>% group_by(Sport) %>% summarise(count=n())
x$Sport <- factor(x$Sport, levels = x$Sport[order(-x$count)])
ggplot(aes(x=Sport, y=count), data=x)+
  geom_bar(width = 0.2, color='#005050', stat="identity")+
  scale_y_log10()+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))+
  ylab("count (log10 scaled)")

```

In this graph we can see the paticipation of athletes in Sports. Aeronautics has
the least number of participants ,only 1 and Athletics has maximum number of 
participants.


### Medals

We have three type of medals Gold, Silver and Bronze, If an athlete did not win
medal then DNW

I am going to plot only the Medal data when the athlete won

```{r}
data$Won <- ifelse(data$Medal!='DNW', 1, 0)
data$Medal <- factor(data$Medal, levels = c('Gold', 'Silver', 'Bronze', 'DNW'))
sum(data$Won)
ggplot(aes(x=Won), data=data)+
  geom_bar(fill='#205050') 
summary(data$Medal)
```

The number of Gold, Silver and Bronze medals are not same, there is a difference 
of a few 100s.

## Bivariant Analysis

In this session we are going to take a look at relation between two variables

### Number of Male and Female athlete over the years

We are going to plot how the number of male and female athlete participation 
changed over the year

```{r}

```



Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
 